Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-08-11 isolation_by_coalescence.tex
when are coal and commute equal

\begin{abstract}
\end{abstract}

@@

The idea of ``resistance'' is an important tool in the lanscape genomics toolbox.
Introduced by \citet{mcrae},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{resistance}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain.
Both electrical resistance and commute time between two nodes
average over all possible paths through the network between the two.
For this reason, it is thought that if genetic distances between samples
from different parts of the landscape
are significantly correlated with the effective resistances between those locations,
with local conductances assigned based on the values of some landscape variable,
then that landscape variable is likely a good indicator of the regions of the map 
where gene flow occurs.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is set equal to its elevation,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more at higher elevations,
perhaps along ridgelines.
\plr{reference some examples}
Other methods \citep{eems} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

@@

We apply the method to the desert tortoise \plr{say more here}.

\plr{I didn't find myself wanting to explain isolation by distance here:
    I don't think we need to? Do you agree?}

@@

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
Therefore, $H$ uniquely determines $G$.

@@

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra.
If $\gamma$ is given, then solving for $G$ is an example of a XXX equation \citep{matrixeqn}.
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZH$ is antisymmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $Z$
so that $G + Z$ remains the generator of a valid Markov chain,
but this is not guaranteed in general.

@@

\subsection*{The simplest example}

@@

@@

@@

Let $\tau_{j} = \inf\{t \ge 0 : X_t = j\}$ and $H_{ij} = \E[\tau_j \,|\, X_0 = i]$.
Then we know that
$$
    (G H)_{ij} = -1 \qquad \text{for} \; i \neq j ,
$$
i.e.,
$$
    GH = - 1 1^T + \diag(x).
$$

What is $x$?  Well, note that
$$
    G = (\diag(x) - 1 1^T) H^{-1}
$$
and so
$$ \begin{aligned}
    0 &= G1 \\
    &= (\diag(x) - 1 1^T) H^{-1} x 
\end{aligned} $$
and so
$$
    x_i = \frac{ 1^T H^{-1} 1 }{ (H^{-1} 1)_i } .
$$
Furthermore,
the Random Target Lemma \citep{aldous} % https://www.stat.berkeley.edu/users/aldous/RWG/Book_Ralph/Ch2.S2.html
tells us that $H \pi = 1$, so thus $H^{-1} 1 = \pi$, the stationary distribution of the chain.
Therefore, $x_i = 1/\pi_i$,
and in fact $1^T H^{-1} 1 = 1$.

In summary,
$$
    G = (\diag(1/H^{-1} 1) - 1 1^T) H^{-1} .
$$

>>>>

\begin{abstract}
\plr{abstract needs editing}
A number of methods commonly used in landscape genetics
use analogies to electrical resistance on a network
to describe and fit barriers to movement across the landscape
using genetic distance data.
These are motivated by a mathematical eqiuvalence of effective resistance
to the mean commute time of a random walk.
Here, we describe the differences between commute time
and coalescence time, which more accurately models genetic distances,
and explore the consequences for inference.
We implement a Bayesian method to do inference under both these models,
and find that inference using commute times
when the data derive from coalescence times
can dramatically mislead in the presence of biased gene flow.
We apply the methods to empirical data,
and find evidence for strongly biased gene flow.
\end{abstract}

@@

The idea of ``resistance'' is an important tool in the lanscape genomics toolbox.
Introduced by \citet{mcrae},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{resistance}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain \citep{levin}.
Both electrical resistance and commute time between two nodes
average over all possible paths through the network between the two.
For this reason, it is thought that if genetic distances between samples
from different parts of the landscape
are significantly correlated with the effective resistances between those locations,
with local conductances assigned based on the values of some landscape variable,
then that landscape variable is likely a good indicator of the regions of the map 
where gene flow occurs.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is set equal to its elevation,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more at higher elevations,
perhaps along ridgelines.
\plr{reference some examples}
Other methods \citep{eems} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

@@

% \plr{I didn't find myself wanting to explain isolation by distance here:
%     I don't think we need to? Do you agree?}

@@

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
Therefore, $H$ uniquely determines $G$.

@@

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{sylvester_eqn}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZH$ is antisymmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $Z$
so that $G + Z$ remains the generator of a valid Markov chain,
but this is not guaranteed in general.

@@

\paragraph{When are coalescence and commute times equal?}
If in a given biological situation the two times were equal --
i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
then inference with either model would be equivalent.
We have already shown that this is not the case in general,
but it turns out that it \emph{is} true
under some restrictive assumptions not uncommonly found in population genetics.
We show in Appendix \ref{sec:com_eq_coal} that
if hitting times are symmetric ($H = H^T$),
then this does occur,
and so the two methods are equivalent for symmetric island models
(where the populations are arranged on a ring 
and migration rates only depend on the distance between them).
Hitting times are not symmetric for a square grid with uniform migration
(the mean time to hit the center from a corner is les than the reverse),
but the square grid is quite close to the torus, which does have symmetric hitting times.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{The simplest example}

@@

Although commute and coalescence times can be quite different,
we show that they are equal (with a particular choice of local diversity)
if populations are arranged on a ring or torus
with symmetric migration.
The similarity of these models to rectangular grids
may explain why in other work, commute time methods have shown relatively good fit to data
simulated with symmetric migration rates.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}

Thanks to David Levin for useful suggestions regarding hitting time calculations,
to Paul Marjoram for useful comments,
and to Brad Shaffer and Evan McCartney-Melstad for useful discussions
and providing the genetic distance data for desert tortoises.
We are also indebted to Chava Weissman, Bridgette Hagerty, Fran Sandmeier, Dick Tracy,
and a large number of field workers for collecting the original tortoise tissue samples.
\plr{also John, Hossein, Desi, ??}

@@

Let $\tau_{j} = \inf\{t \ge 0 : X_t = j\}$ and $H_{ij} = \E[\tau_j \,|\, X_0 = i]$.
Then we know that
$$
    (G H)_{ij} = -1 \qquad \text{for} \; i \neq j ,
$$
i.e.,
\begin{align} \label{eqn:GH}
    GH = - 1 1^T + \diag(x).
\end{align}

What is $x$?  Well, note that
$$
    G = (\diag(x) - 1 1^T) H^{-1}
$$
and so
$$ \begin{aligned}
    0 &= G1 \\
    &= (\diag(x) - 1 1^T) H^{-1} x 
\end{aligned} $$
and so
$$
    x_i = \frac{ 1^T H^{-1} 1 }{ (H^{-1} 1)_i } .
$$
Furthermore,
the Random Target Lemma \citep{aldous} % https://www.stat.berkeley.edu/users/aldous/RWG/Book_Ralph/Ch2.S2.html
tells us that if $\pi$ is the stationary distribution of the chain, then 
$(H \pi)_i$ does not depend on $i$,
i.e., that $H \pi \propto 1$,
and thus if $H^{-1} 1 = \nu$, 
then $\pi = \nu / \sum_i \nu_i$.
This says exactly that $x_i = 1/\pi_i$.

In summary, $\pi = H^{-1} 1 / 1^T H^{-1} 1$, and 
\begin{align}
    G 
    &= (\diag(1/\pi) - 1 1^T) H^{-1} .
\end{align}
Note that this equation only specifies $H$ up to a constant added to each column:
i.e., if given $G$ one obtains a matrix $Y$ solving $GY = \diag(1/\pi) - 1 1^T$,
then $H_{ij} = Y_{ij} - Y_{jj}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Equality of commute and coalescence times}
\label{sec:com_eq_coal}

Under what conditions are coalescence times and commute times (with some local diversity values) equal?
In other words, for what choices of $G$, $\gamma$, and $q$ does
\begin{align}
    \comdist = (H + H^T)/4 + (q 1^T + 1 q^T)/2
\end{align}
solve equation \eqref{eqn:C_matrix}?
Writing this out, this says that
\begin{align} \label{eqn:R_coal}
    ( GH + (GH)^T + GH^T + (GH^T)^T )/4 + (Gq 1^T + 1 (Gq)^T)/2
    &=
    \diag(q) \diag(\gamma) - 1 1^T .
\end{align}
(To simplify this we used the fact that $\diag(H) = 0$ and $G1 = 0$.)

It is not clear what can be said about the general case because of the presence of $HG^T$,
but if we assume that $H$ is symmetric, we can make progress,
because then we have that $GH = GH^T$, 
and so equation \eqref{eqn:GH} says that all four terms in the first group are equal,
and equation \eqref{eqn:R_coal} simplifies to
\begin{align} \label{eqn:R_eq_C}
    \diag(1/\pi) + (Gq 1^T + 1 (Gq)^T)/2
    &=
    \diag(q) \diag(\gamma) ,
\end{align}
where the $-1 1^T$ terms on each side canceled.
Because $G1=0$,
the most obvious solution to this is if $q = c 1$ and $c \gamma = 1/\pi$, for some constant $c$
(although there is a broader family of solutions).

In summary, if hitting times are symmetric
and coalescence rates are equal to the inverse of the stationary distribution,
then coalescence times are equal to commute times plus one.
This is fairly restrictive, 
but does occur if the population configuration is isotropic
(as for instance in an all-connected-to-all island model with equal migration rates)
or in populations arranged around a ring
with migration rates depending only on the distance between them.

Can we solve equation \eqref{eqn:R_eq_C} more generally?
For that equation to hold, we need $Gq 1^T +  (Gq)^T$ to be diagonal,
i.e., that $(Gq)_i + (Gq)_j = 0$ for all $i \neq j$.
For any $k \neq \ell$, there exists a vector $u^{(k\ell)}$ such that $(Gu^{(k\ell)})_i = \delta_{ik} - \delta_{i\ell}$;
the only possible $q$ for which $Gq 1^T +  (Gq)^T$ is diagonal
are of the form $q = \alpha 1 + \beta u^{(k,\ell)}$ for some $k \neq \ell$ and some constants $\alpha$ and $\beta$.
We would then need
\begin{align}
    1/\pi_i + \delta_{ik} - \delta_{i\ell} = q_i \gamma_i \qquad \text{for all} i.
\end{align}
This implies that for any Markov chain with symmetric hitting times,
we can find diversity values ($q$) and coalescence rates ($\gamma$) that make $\comdist = C$
(and in fact there are many ways to do this).
However, there is not a general solution for $q$ if coalescence rates are also given (as is the case in practice).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Something else?}

\citet{strobeck1987average} showed that $\diag(C)$ was constant and does not depend on movement rates
for any isotropic conservative migration model.
Can we explain this?
(Also cite earlier.)
